---
author: "Kasper D. Hansen"
title: "BSgenome - Views"
---

```{r front, child="front.Rmd", echo=FALSE}
```

## Dependencies

This document has the following dependencies:

```{r dependencies, warning=FALSE, message=FALSE}
library(BSgenome)
library(BSgenome.Scerevisiae.UCSC.sacCer2)
library(AnnotationHub)
```

Use the following commands to install these packages in R.

```{r biocLite, eval=FALSE}
source("http://www.bioconductor.org/biocLite.R")
biocLite(c("BSgenome",
           "BSgenome.Scerevisiae.UCSC.sacCer2", "AnnotationHub"))
```

## Overview

We continue our treatment of `r Biocpkg("Biostrings")` and `r Biocpkg("BSgenome")`

## Other Resources

## Views

`Views` are used when you have a single big object (think chromosome or other massive dataset) and you need to deal with (many) subsets of this object.  `Views` are not restricted to genome sequences; we will discuss `Views` on other types of objects in a different session.

Technically, a `Views` is like an `IRanges` couple with a pointer to the massive object.  The `IRanges` contains the indexes.  Let's look at `matchPattern` again:
```{r views1}
library("BSgenome.Scerevisiae.UCSC.sacCer2")
dnaseq <- DNAString("ACGTACGT")
vi <- matchPattern(dnaseq, Scerevisiae$chrI)
vi
```
We can get the `IRanges` component by
```{r views2}
ranges(vi)
```
The `IRanges` gives us indexes into the underlying subject (here chromosome I).  To be clear, compare these two:
```{r views3}
vi
Scerevisiae$chrI[ start(vi):end(vi) ]
```
The `Views` object also look a bit like a `DNAStringSet`; we can do things like
```{r views4}
alphabetFrequency(vi)
```
The advantage of `Views` is that they don't duplicate the sequence information from the subject; all they keep track of are indexes into the subject (stored as `IRanges`).  This makes it very (1) fast, (2) low-memory and makes it possible to do things like
```{r views5}
shift(vi, 10)
```
where we now get the sequence 10 bases next to the original match.  This could not be done if all we had were the bases of the original subsequence.

`Views` are especially powerful when there are many of them.  A usecase I often have are the set of all exons (or promoters) of all genes in the genome.  You can use `GRanges` as `Views` as well.  Lets look at the hits from `vmatchPattern`.

```{r viewsVMatchPattern}
gr <- vmatchPattern(dnaseq, Scerevisiae)
vi2 <- Views(Scerevisiae, gr)
```

Now, let us do something with this.  First let us get gene coordinates from `r Biocpkg("AnnotationHub")`.

```{r annotationHub}
ahub <- AnnotationHub()
qh <- query(ahub, c("sacCer2", "genes"))
qh
genes <- qh[[which(qh$title == "SGD Genes")]]
genes
```

Let us compute the GC content of all promoters in the yeast genome.
```{r promoterGCcontent}
prom <- promoters(genes)
head(prom, n = 3)
```
We get a `warning` that some of these promoters are out-of-band (see the the second and third element in the `prom` object; they have negative values for their ranges).  We clean it up and continue
```{r promoterGCcontent2}
prom <- trim(prom)
promViews <- Views(Scerevisiae, prom)
gcProm <- letterFrequency(promViews, "GC", as.prob = TRUE)
head(gcProm)
```
In the previous `r Biocpkg("Biostrings")` session we computed the GC content of the yeast genome.  Let us do it again, briefly
```{r genomeGC}
params <- new("BSParams", X = Scerevisiae, FUN = letterFrequency, simplify = TRUE)
gccontent <- bsapply(params, letters = "GC")
gcPercentage <- sum(gccontent) / sum(seqlengths(Scerevisiae))
gcPercentage
```
Let us compare this genome percentage to the distribution of GC content for promoters
```{r plotGC, fig=TRUE}
plot(density(gcProm))
abline(v = gcPercentage, col = "red")
```
At first glance, the GC content of the promoters is not very different from the genome-wide GC content (perhaps shifted a bit to the right).

```{r back, child="back.Rmd", echo=FALSE}
```

