---
title: "SummarizedExperiment"
author: "Kasper D. Hansen"
---

```{r front, child="front.Rmd", echo=FALSE}
```

## Dependencies

This document has the following dependencies:

```{r dependencies, warning=FALSE, message=FALSE}
library(SummarizedExperiment)
library(GenomicRanges)
library(airway)
```

Use the following commands to install these packages in R.

```{r biocLite, eval=FALSE}
source("http://www.bioconductor.org/biocLite.R")
biocLite(c("SummarizedExperiment", "GenomicRanges", "airway"))
```

## Overview

We will present the `SummarizedExperiment` class from `r Biocpkg("GenomicRanges")` package; an extension of the `ExpressionSet` class to include `GRanges`.
  
This class is suitable for storing processed data particularly from high-throughout sequencing assays.
  
## Start

An example dataset, stored as a `SummarizedExperiment` is available in the `r Biocexptpkg("airway")` package.  This data represents an RNA sequencing experiment, but we will use it only for illustrating the class.

```{r airway}
library(airway)
data(airway)
airway
```

This looks similar to - and yet different from - the `ExpressionSet`.  Things now have different names, representing that these classes were designed about 10 years apart.

We have 8 samples and 64102 features (genes).

Some aspects of the object are very similar to `ExpressionSet`, although with slightly different names and types:

`colData` contains phenotype (sample) information, like `pData` for `ExpressionSet`. It returns a `DataFrame` instead of a `data.frame`:
```{r colData}
colData(airway)
```
You can still use `$` to get a particular column:
```{r getColumn}
airway$cell
```
`metadata` is like `experimentData` from `ExpressionSet`.  This slot is often un-used; this is the case for this object:
```{r exptData}
metadata(airway)
```
`colnames` are like `sampleNames` from `ExpressionSet`; `rownames` are like `featureNames`.
```{r names}
colnames(airway)
head(rownames(airway))
```

The measurement data are accessed by `assay` and `assays`.  A `SummarizedExperiment` can contain multiple measurement matrices (all of the same dimension).  You get all of them by `assays` and you select a particular one by `assay(OBJECT, NAME)` where you can see the names when you print the object or by using `assayNames`. In this case there is a single matrix called `counts`:
```{r assay}
airway
assayNames(airway)
assays(airway)
head(assay(airway, "counts"))
```

So far, this is all information which could be stored in an `ExpressionSet`.  The new thing is that `SummarizedExperiment` allows for a `rowRanges` (or `granges`) data representing the different features.  The idea is that these `GRanges` tells us which part of the genome is summarized for a particular feature.  Let us take a look
```{r rowRanges}
length(rowRanges(airway))
dim(airway)
rowRanges(airway)
```
See how `rowRanges` is a `GRangesList` (it could also be a single `GRanges`).  Each element of the list represents a feature and the `GRanges` of the feature tells us the coordinates of the exons in the gene (or transcript).  Because these are genes, for each `GRanges`, all the ranges should have the same `strand` and `seqnames`, but that is not enforced.

In total we have around 64k "genes" or "transcripts" and around 745k different exons.
```{r numberOfExons}
length(rowRanges(airway))
sum(elementLengths(rowRanges(airway)))
```

For some operations, you don't need to use `rowRanges` first, you can use the operation directly on the object.  Here is an example with `start()`:
```{r start}
start(rowRanges(airway))
start(airway)
```
You can use `granges()` as synonymous for `rowRanges()`.

Subsetting works like `ExpressionSet`: there are two dimensions, the first dimension is features (genes) and the second dimension is samples.

Because the `SummarizedExperiment` contains a `GRanges[List]` you can also use `subsetByOverlaps`, like
```{r subsetByOverlaps}
gr <- GRanges(seqnames = "1", ranges = IRanges(start = 1, end = 10^7))
subsetByOverlaps(airway, gr)
```

```{r back, child="back.Rmd", echo=FALSE}
```
